---
title: "Growth Standards"
output:
 rmarkdown::html_vignette:
   toc: true
   number_sections: true
bibliography: references.bib
vignette: >
 %\VignetteIndexEntry{Growth Standards}
 %\VignetteEngine{knitr::rmarkdown}
 %\VignetteEncoding{UTF-8}
---


```{r label = "setup", include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, fig.align = "center")
library(pedbp)
```


# CDC Growth Charts

Using the [Percentile Data Files with LMS values](https://www.cdc.gov/growthcharts/percentile_data_files.htm)
provided by the CDC, we provide eight distribution tables which have been
combinded into seven.  A

| CDC Source                    | pedbp metric               |
| :---------------------------- | :------------------------- |
| BMI for age                   | bmi_for_age                |
| head circumference for age    | head_circumference_for_age |
| length for age for infants    | length_for_age             |
| stature for age               | height_for_age             |
| weight for age for infants    | weight_for_age             |
| weight for age                | weight_for_age             |
| weight for length for infants | weight_for_length          |
| weight for stature            | weight_for_height          |

All lengths/heights are in centimeters, ages in months, and weights in
kilograms.

The length-for-age and height-for-age methods were needed for the blood
pressure methods.

# WHO Growth Charts

# LMS

All methods use the published LMS parameters to define z-scores,
percentiles, and quantiles for skewed distributions.  L is a $\lambda$
parameter, the Box-Cox transformation power; $M$ the median value, and $S$ a
generalized coefficient of variation.  For a given percentile or z-score, the
corresponding physical measurement, $X,$ is defined as

$$X = \begin{cases}
   M \left(1 + \lambda S Z \right)^{\frac{1}{\lambda}} & \lambda \neq 0 \\
   M \exp\left( S Z \right) & \lambda = 0.
\end{cases}$$

From this we can get the z-score for a given measurement $X:$

$$ Z = \begin{cases}
  \frac{\left(\frac{X}{M}\right)^{\lambda} - 1}{\lambda S} & \lambda \neq 0 \\
  \frac{\log\left(\frac{X}{M}\right) }{ S } & \lambda = 0.
  \end{cases}$$

Percentiles are determined using the standard normal distribution of z-scores.

For all eight of the noted methods we provide a distribution function,
quantile function, and function that returns z-scores.

# Metrics

## Height for Age

A height for age chart based on the CDC data:

```{r echo = FALSE}
lfa <- data.table::CJ(p = c(0.01, 0.05, 0.10, 0.25, 0.50, 0.75, 0.90, 0.95, 0.99),
                      age = seq(0, 18*12, by = 3),
                      male = 0:1)

lfa[age <  36, l := q_height_for_age(p = p, age = age, male = male), by = .(p, age, male)]
lfa[age >= 36, l := q_height_for_age(p = p, age = age, male = male), by = .(p, age, male)]
lfa[, lab := paste(p * 100, "%")]
lfa[, male := factor(male, 0:1, c("Female", "Male"))]
percentile_factor <- function(p) {
  factor(p, levels = sort(unique(p)), labels = paste0(sort(unique(p)) * 100, "th"))
}
lfa[, p := percentile_factor(p)]

g <- function(lfa) {
  ggplot2::ggplot(lfa) +
    ggplot2::theme_bw() +
    ggplot2::aes(x = age, y = l, color = p) +
    ggplot2::geom_line() +
    ggplot2::facet_wrap( ~ male) +
    ggplot2::scale_x_continuous(name = "Age"
                                , breaks = seq(0, max(lfa$age) + 12, by = 12)
                                , labels = paste(
                                                   paste0(seq(0, max(lfa$age) + 12, by = 12), "m")
                                                 , paste0(seq(0, max(lfa$age) + 12, by = 12) / 12, "yr")
                                                 , sep = "\n")
                                ) +
    ggplot2::scale_y_continuous(name = "Length or Height (cm)") +
    ggplot2::scale_color_hue(name = "Length for Age percentile") +
    ggplot2::theme(
                   legend.position = "bottom"
                   )
}
```




```{r label = "lfa_female", echo = FALSE, fig.width = 8, fig.height = 6}
g(lfa[male == "Female"])
```




```{r label = "lfa_male", echo = FALSE, fig.width = 8, fig.height = 6}
g(lfa[male == "Male"])
```


## Weight for Age

There are two methods for determining weight for age both based on CDC
National Center for Health Statistics data: one for infants (weighed laying
flat) up to 36 months [@cdc_weight_for_age_0_36],
and one for children (weighed on a standing scale) over 24 months
[@cdc_weight_for_age].

A 33 pound (
`r  round(33 * 0.453592, 2) `
kg) 4 year old male is in the
`r  paste0(round(100 * p_weight_for_age(33 * 0.453592, age = 4 * 12, male = 1), 2), 'th')  `
percentile.

```{r}
p_weight_for_age(33 * 0.453592, age = 4 * 12, male = 1)
```


The 20th percentile weight for an 18 month old infant female is
`r  round(q_weight_for_age(p = 0.2, age = 18, male = 0), 5) `
kg.

```{r}
round(q_weight_for_age(p = 0.2, age = 18, male = 0), 5)
```






# References
<div id="refs"></div>

# Session Info

```{r label = "sessioninfo"}
sessionInfo()
```

